---
title: "2 Calculation Exercises"
author: ''
date: ''
output:
  html_document: default
---

```{r,echo=FALSE}
options(digits=12)
unit.test.points=0
```

# General instructions.

Exercise 1 is required and must be solved using both R and SAS. This exercise is worht 20 points, 10 points each for R and SAS. The R portion is included in this document and includes a unit test that checks the results of your calculations. 

Exercises 2 is also required, but you may choose to use either R or SAS. Make that you include in your code comments documenting your choice, don't simply submit the results without explanation. This exercise is worth 10 points.

Exercises 3-5 are optional, choose two of three and document your choice. Your are also allowed to choose from either R or SAS. Choose only 1, each will be worth 10 points, and the entire assignment will be worth 50 points.

Be sure to document your code and comment on the results.

# Exercise 1 (R and SAS, required)


Calculate Cohen's $d$ for calories per recipe, calories per serving and servings per recipe, comparing years 1936 and 2006. Use the formula
$$
d = \frac{|m_1-m_2|}{\sqrt{(s_1^2 + s_2^2)/2}}
$$

## Part a. 
Enter the R code in the code chunk below, and assign a value to `d_12` for unit testing.

```{r}
# Variables for "calories per recipe" (CPR) to be input into formula
 m_1 <- 2123.8 #Mean CPR 1936
 m_2 <- 3051.9 #Mean CPR 2006
 s_1 <- 1050.0 #SD CPR 1936
 s_2 <- 1496.2 #SD CPR 2006

# Assigning formula to variable d_12 for next chunk.
 d_12<- (abs(m_1-m_2))/(sqrt((s_1^2 + s_2^2)/2))
print(d_12);
# Result: 0.718065555553

# Variables for "calories per serving" (CPS) to be input into formula
m_1 <- 268.1 #Mean CPS 1936
m_2 <- 285.6 #Mean CPs 2006
s_1 <- 124.8 #SD CPS 1936
s_2 <- 118.3 #SD CPS 2006

# Assigning formula to variable d_12 for next chunk.
d_12 <- (abs(m_1-m_2))/(sqrt((s_1^2 + s_2^2)/2))
print(d_12)
# Result: 0.143922236212

# Variables for "servings per recipe" (SPR) to be input into formula
m_1 <- 12.9 #Mean SPR 1936
m_2 <- 12.4 #Mean SPR 2006
s_1 <- 13.3 #SD SPR 1936
s_2 <- 14.3 #SD SPR 2006

# Assigning formula to variable d_12 for next chunk.
d_12 <- (abs(m_1-m_2))/(sqrt((s_1^2 + s_2^2)/2))
print(d_12)
# Result: 0.0362081257545



```


```{r, unit test for exercise 1,echo=FALSE}


if(abs(d_12-0.7181)<0.0001) {
  print(unit.test.points <- unit.test.points+8)
} else {
  cat('d is not assigned the correct value','\n')
}
#CPR Results: 8
#CPS Result: d is not assigned the correct value
#SPR Result: d is not assigned the correct value
```

## Part b.
For the SAS portion, you may use PROC IML or macro language. Use `2 Calculations Template.sas` to start. This file contains unit tests for Exercises 1 and 2.

# Exercise 2 (R or SAS, required)

The probablity of an observation $x$, when taken from a normal population with mean $\mu$ and variance $\sigma^2$ is calculated by
$$
L (x ; \mu, \sigma^2) = \frac{1}{\sigma \sqrt{2 \pi}^{}} e^{- \frac{(x - \mu)^2}{2 \sigma^2}}
$$
For values of $x = \{ 0.1, 0.2 \}$, write code to calculate $L (x ; \mu = 0, \sigma = 1)$. Assign the values to `l_1` and `l_2`.
  
### Answer

If you choose to do this exercise in R, use the unit test below.

```{r}
l_1 <- NA
l_2 <- NA
```

### Unit Test (R)
```{r, unit test for exercise 2,echo=FALSE}
if(abs(l_1 - 0.396952)<1e-6) {
  print(unit.test.points <- unit.test.points + 4)
} else {
  print("likelihood for x = 0.1 incorrect")
}

if(abs(l_2 - 0.391043)<1e-6 ) {
  print(unit.test.points <- unit.test.points + 4)
} else {
  print("likelihood for x = 0.2 incorrect")
}
```

If you choose to do this exercise in SAS, you must do this in PROC IML, using the unit test template `2 Calculations.sas`.

# Exercise 3 (R or SAS)

## Part a

Write code to compute
  
  $$7 - 1 \times 0 + 3 \div 3$$
  
Type this in verbatim, using only numbers, `-`,`*` and `/`, with no parenthesis. Do you agree with the result? Explain why, one or two sentences.

#### Answer

```{r}

```

## Part b

  According to "Why Did 74% of Facebook Users Get This Wrong?" (<http://www.classroomprofessor.com/teaching-math/why-did-74pc-of-facebook-users-get-this-wrong/>), most people would compute the result as 1. 
  Use parenthesis `( )` to produce this result.

#### Answer

```{r}

```

## Part c

Several respondents to the survey cited in Part 2 gave the answer 6. Add *one* set of parenthesis to produce this result.

#### Answer
```{r}

```

# Exercise 4. (R or SAS)

### Part a

Quoting from Wansink and Payne

> Because of changes in ingredients,
the mean average calories in a recipe increased by 928.1 (from
2123.8 calories ... to 3051.9 calories
... ), representing a 43.7% increase.

Show how 43.7% is calculated from 2123.8 and 3051.9, and confirm W&P result.

> The resulting increase of 168.8 calories (from 268.1 calories ... to 436.9 calories ...) represents a 63.0% increase ... in calories per serving.

### Part b

Repeat the calculations from above and confirm the reported 63.0% increase in calories per serving.

```{r}
```

### Part c

Using values from Table 1 Wansink, calculate the percent change in calories per serving from 1997 to 2006. How does this value compare to 63.0 increase as quoted in part b.?


```{r}
```


# Exercise 5. (R or SAS)

From Wansink and Payne
> Given that the average 2006 recipe had 1.1 fewer servings than in 1936, the average calorie density per serving size has increased by 37.4% 

From https://foodpsychology.cornell.edu/research/joy-cooking-too-much-70-years-calorie-increases-classic-recipes
> "Also, the mean average calories per serving increased in 17 out of the 18 recipes by 37.4% from 268.1 calories to 436.9 calories"

https://foodpsychology.cornell.edu/discoveries/joy-cooking-too-much
>This expanded portion size helps explain why calories per serving have increased from an average of 168.8 calories to 436.9 calories, which is a 63% increase in calories per serving

Show calculations in R or SAS to either confirm these statements (using values from Wansink Table 1) or show where the quoted values are inconsistent with Wansink Table 1.


# Total points from unit tests
```{r}
unit.test.points
```



