---
title: "Joy Of Cooking Midterm Project Prep"
author: "Peter Claussen"
date: "6/25/2018"
output:
  html_document: default
---

# Part 1 - Assign Recipes

First, generate a list of all files in the Recipes folder.
```{r,eval=FALSE}
recipes <- list.files(path = "../data/Recipes", pattern = '*.pdf')
```

Make this reproducible.
```{r,eval=FALSE}
set.seed(10010)
```

Create two randomly sampled list of recipe files and cycle this list through 80 rows

```{r,eval=FALSE}
recipe.pairs <- matrix(c(sample(recipes),sample(recipes)),ncol=2,nrow=80)
```

Are there any pairs across rows?

```{r,eval=FALSE}
sum(recipe.pairs[,1]==recipe.pairs[,2])
```

Write to a table

```{r,eval=FALSE}
write.csv(recipe.pairs, file="Recipe.Table.csv",row.names = FALSE)
```

Read the table in and print:

```{r}
Recipe.Table <- read.csv("Recipe.Table.csv",header=FALSE,skip=1)
print(Recipe.Table)
```

# Part 2 - Check Nutrient Data.

We will be using the USDA food nutrient database, found at https://www.ars.usda.gov/northeast-area/beltsville-md-bhnrc/beltsville-human-nutrition-research-center/nutrient-data-laboratory/docs/usda-national-nutrient-database-for-standard-reference/

Run this to make sure your system can read the required nutrient database tables.
```{r}
weight.dat <- read.table("../data/SR-Leg_ASC/WEIGHT.txt",header=FALSE,sep="^",quote="~")
names(weight.dat) <- c("NDB_No","Seq","Amount","Msre_Desc","Gm_Wgt","Num_Data_Pts","Std_Dev")
head(weight.dat)
summary(weight.dat)
```

```{r}
food_des.dat <- read.table("../data/SR-Leg_ASC/FOOD_DES.txt",header=FALSE,sep="^",quote="~")
names(food_des.dat) <- c("NDB_No","FdGrp_Cd","Long_Desc","Shrt_Desc",
"ComName","ManufacName","Survey","Ref_desc","Refuse","SciName",
"N_Factor","Pro_Factor","Fat_Factor","CHO_Factor")
dim(food_des.dat)
head(food_des.dat)
summary(food_des.dat)
```

```{r}
nut_data.dat <- read.table("../data/SR-Leg_ASC/NUT_DATA.txt",header=FALSE,sep="^",quote="~")
names(nut_data.dat) <- c("NDB_No","Nutr_No","Nutr_Val","Num_Data_Pts","Std_Error",
"Src_Cd","Deriv_Cd","Ref_NDB_No","Add_Nutr_Mark","Num_Studies",
"Min","Max","DF","Low_EB","Up_EB","Stat_cmt","AddMod_Date")
head(nut_data.dat)
summary(nut_data.dat)
```
